### 2022-11-28 response-letter.md

Dear editor,

Thank you for the inviting a specialist on large scale biodiversity patterns of
Africa to review our manuscript. We were happy to see that the first reveiwer
was satisfied with our response, and that the second reviewer also saw value in
our contribution. The new comments were very pertinent and we hope that this
letter and the changes we've made to the paper are satisfactory.

We have improved the text in order to avoid misleading phrasing. For example, we
understand that using the words "updated ranges" can be read as distribution
modelling, which is not the case: here, we simply use a method to identify areas
where the predator range does not match with their prey's range, which might
indicate distribution uncertainty areas.

We have also addressed reviewer's concerns about the GBIF data (see below): our
investigations show that most of the data we were using are recent (from the
last 20 years). We decided to add a threshold on the year 2000 and restrict the
time extension of our query the same way we do for the geographical extension -
not as a filtering step in terms of Ecology, but of likelyhood of getting the
most accurate data. Any refined filtering other than that would hinder the goals
of our paper. Finally, we have improved the methods and the discussion sections
to include more detailed descriptions of the data being used and to give more
context to the results. We detail specific changes below (line numbers refer to
the clean, revised PDF file).

@@

That being said, the foodweb database that we chose is known for being
representative for these animals on the whole African continent (as we can see,
for example, in de Visser, Freymann and Olff, 2011). Although our dataset miss a
few links, again, it was not our goal to make a comprehensive analysis of all
the area where a predator would miss a prey based on all the knowledge available
or not, but rather demonstrate what can happen if we deal with incomplete
datasets, highlight the importance of accounting for uncertainty when dealing
with species distribution data, and call to attention the need to make more
interaction data open. Moreover, many of the prey items on Baskerville's dataset
were also found outside the Serengeti park area, and our method only needed one
link connecting a predator to a herbivore to "keep" a pixel from a predator's
range. Therefore, the impact of the missing links on the foodweb should be of
great magnitude to invalidate our methods.

Finally, some of the results we highlight would not change with the addition of
new prey items. For example, the *Canis aureus* results are due mainly to an
apparent taxonomic disagreement - which is indeed connected to its interaction
data, but this would not change with a more complete list of preys for this
species.

To emphasize and clarify these points, we have changed lines 83-86 and 171-175 on
the manuscript.

@@

>*"4. The gbif records for large mammals are especially problematic. Did the
>authors include observations as well? iNaturalist for instance now feeds data
>directly to GBIF. Having museum specimens of lions and other large/medium
>mammals are especially problematic and difficult to achieve nowadays. Many
>countries such as Malawi and Mozambique conduct very little research and many
>areas are not documented, not even on GBIF. Therefore the occurrences we see on
>GBIF are biased towards countries that have available data. How did the authors
>account for this? See https://doi.org/10.1093/sysbio/syaa090"*

Similar to the questions above, this one is also about incompleteness of data.
As we stated on our previous review letter, "we didn't perform any [...] data
cleaning procedures because this could hinder the goals of the paper, which was
to investigate the characteristics of the data available for research. As we
needed to compare two sets of raw data, and data cleaning protocols are
dependent on the use objectives, we chose to avoid data manipulation both for
GBIF and IUCN datasets". For that reason, our query for the GBIF data did not
include any filters other than the ones described on the paper for the reasons
listed above.

@@

Additionally, we could add a plethora of other datasets both on
interactions and occurrences, but we decided to use only the IUCN, the GBIF and
the Baskerville's data to keep the analysis simple (although not simplistic),
straightforward, and reflective of the volume of data generally used by
ecologists.

Finally, the fact that many data are still to be made open is precisely the
major motivation of this paper. As computational ecologists and open data
creators/managers/advocates, we are very aware of the biases produced by
undocumented data. We expect to call attention about this issue with our proof
of concept, as it demonstrates that we gain a lot of uncertainty when we miss a
link because of an information that is not accessible. The implications of this
complex issue for conservation are multifold, and we believe that more open data
is one of the possible solutions.

>*"2.    The resolution: it is well studied that IUCN range maps should not be
>converted to grids smaller than 50km due to commission errors. This becomes a
>major problem when dealing with widespread species such as the ones used in
>this study: African large and medium mammals. See
>https://doi.org/10.1111/1365-2664.12771 and
>https://doi.org/10.1073/pnas.070446910"*

We agree and we have updated our rasters to be now at 50km of resolution. Our
overall results did not change with that (see line 125 and Results section,
especially Table 1).

@@

>*"3. Gbif records: using records from any period in time from gbif will include
>historical relics from extinct populations. Many large and medium mammals,
>especially large carnivores have suffered major declines in the last decades in
>Africa. How did the authors deal with the fact that records from a species
>recorded 50 years ago may represent extinct populations?"*

This is a very interesting comment and we have explored the temporal
distribution of the GBIF records we have retrieved. As you can see below, most
of the data is recent, from after the year 2000. This means that our previous
results are very likely to represent our current state of biodiversity
(considering a temporal autocorrelation effect). However, we agree that extinct
populations could be a real problem for the validation step of our method, and
therefore we decided to add a temporal restriction to our query. We have then
re-run our analyses with only GBIF occurrences registered after the year 2000.
Still, our results are overall the same. We have updated the absolute values on
lines 212-215, 225-228, 259, 274-275, and 285-289.

![](figures/GBIF_years_prop_nb_2000.png)

### 2022-11-28 response-letter.md

Dear editor,

Thank you for inviting a specialist on large scale biodiversity patterns of
Africa to review our manuscript. We were happy to see that the first reviewer
was satisfied with our response, and that the second reviewer also saw value in
our contribution. The new comments were very pertinent and we hope that this
letter and the changes we've made to the paper are satisfactory.

@@

We have improved the text in order to avoid misleading phrasing. For example, we
understand that using the words "updated ranges" can be read as distribution
modelling, which is not the case: here, we simply use a method to identify areas
where the predator range does not match with their prey's range, which might
indicate distribution uncertainty areas.

We have also addressed the reviewer's concerns about the GBIF data (see below): our
investigations show that most of the data we were using are recent (from the
last 20 years). We decided to add a threshold on the year 2000 and restrict the
time extension of our query the same way we did for the geographical extension -
not as a filtering step in terms of Ecology, but of likelihood of getting the
most accurate data. Any refined filtering other than that would hinder the goals
of our paper. Finally, we have improved the methods and discussion sections
by including more detailed descriptions of the data being used and by giving more
context to the results. We detail specific changes below (line numbers refer to
the clean, revised PDF file).

@@

>*"1. In the extent of the study, the authors mention that they “restricted the
>rasters a spatial extent comprised between latitudes 35°S and 40°N and
>longitudes 20°W and 55°E. “ – which is the whole of Africa. Since the study
>uses a dataset from Tanzania (Serengeti food web dataset), roughly 3% of the
>continent, how did the authors deal with the presence of putative prey items in
>other parts of the continent that are absent from the Serengeti food web
>dataset? For example Papio ursinus, Sylvicapra caffra and S. grimmia don't
>occur in Tanzania but are prey items to many carnivores in Africa."*

>*"5. Taxonomic groups: Carnivores feed on a diversity of items, beyond
>herbivores. Their diet might include birds, fish, reptiles, amphibians, and
>even invertebrates. How did the authors account for the fact that despite
>missing the herbivores from Serengeti, there were other sources of protein that
>could sustain the species in the continent?"*

We have found that these two comments address the same issue: the
representativity of the Serengeti food web. First, we would like to emphasize
that our analysis is meant to be a proof of concept, showing the consequences of
data limitation. Therefore, rather than trying to gather an exhaustive food web
dataset - with information for all possible interactions for all occurring
mammals in the African continent -, we looked for a complete, readily
available, widely used and well known dataset which could represent the main
interactions in that area. Interaction data is difficult to collect, especially
at broad scales. Because of that, we expected that no food web would be complete,
and we are aware of the biases of our interactions data. However, the goal of
this paper is to investigate precisely what happens if we need to deal with
incomplete and biased datasets, because we believe that this might lead us to
describe unrealistic ecological communities at the local scale. More than that,
we believe that data incompleteness is not a limitation of our study, but one of
its strengths, because we present a simple method to identify and describe these
limitations.

That being said, the food web dataset that we chose is known for being
representative for these animals on the whole African continent (as we can see,
for example, in de Visser, Freymann and Olff, 2011). Even though our dataset misses a
few interactions, again, it was not our goal to conduct a comprehensive analysis of all
the areas where a predator would miss a prey based on all the knowledge available. Our goal was rather to demonstrate what can happen if we deal with incomplete
datasets, to highlight the importance of accounting for uncertainty when dealing
with species distribution data, and to call to attention the need to make more
interaction data open. Moreover, many of the prey items on Baskerville's dataset
were also found outside the Serengeti park area, and our method only needed one
link connecting a predator to a herbivore to "keep" a pixel from a predator's
range. Therefore, the impact of missing interactions on the food web should be of
great magnitude to invalidate our method.

@@

Finally, some of the results we highlight would not change with the addition of
new prey items. For example, the *Canis aureus* results are due mainly to an
apparent taxonomic disagreement - which is indeed connected to its interaction
data, but this would not change with a more complete list of preys for this
species.

To emphasize and clarify these points, we have changed lines 83-86 and 171-175 on
the manuscript.

>*"4. The gbif records for large mammals are especially problematic. Did the
>authors include observations as well? iNaturalist for instance now feeds data
>directly to GBIF. Having museum specimens of lions and other large/medium
>mammals are especially problematic and difficult to achieve nowadays. Many
>countries such as Malawi and Mozambique conduct very little research and many
>areas are not documented, not even on GBIF. Therefore the occurrences we see on
>GBIF are biased towards countries that have available data. How did the authors
>account for this? See https://doi.org/10.1093/sysbio/syaa090"*

Similar to the questions above, this one is also about incompleteness of data.
As we stated in our previous review letter, "we didn't perform any [...] data
cleaning procedures because this could hinder the goals of the paper, which was
to investigate the characteristics of the data available for research. As we
needed to compare two sets of raw data, and data cleaning protocols are
dependent on the use objectives, we chose to avoid data manipulation both for
GBIF and IUCN datasets". For that reason, our query for the GBIF data did not
include any filters other than the ones described in the paper for the reasons
listed above.

Additionally, we could have added a plethora of other datasets both on
interactions and occurrences, but we decided to use only the IUCN, the GBIF and
the Baskerville's data to keep the analysis simple (although not simplistic),
straightforward, and reflective of the volume of data generally used by
ecologists.

Finally, the fact that many data are still to be made open is precisely the
major motivation of this paper. As computational ecologists and open data
creators/managers/advocates, we are very aware of the biases produced by
undocumented data. We expect to call attention about this issue with our proof
of concept, as it demonstrates that we gain a lot of uncertainty when we miss interactions because of inaccessible information. The implications of this
complex issue for conservation are multifold, and we believe that more open data
is one of the possible solutions.

@@

>*"2.    The resolution: it is well studied that IUCN range maps should not be
>converted to grids smaller than 50km due to commission errors. This becomes a
>major problem when dealing with widespread species such as the ones used in
>this study: African large and medium mammals. See
>https://doi.org/10.1111/1365-2664.12771 and
>https://doi.org/10.1073/pnas.070446910"*

We agree and we have updated our rasters to be at 50km of resolution. Our
overall results did not change qualitatively after fixing the resolution (see line 125 and Results section,
especially Table 1).

@@

>*"3. Gbif records: using records from any period in time from gbif will include
>historical relics from extinct populations. Many large and medium mammals,
>especially large carnivores have suffered major declines in the last decades in
>Africa. How did the authors deal with the fact that records from a species
>recorded 50 years ago may represent extinct populations?"*

This is a very interesting comment and we have explored the temporal
distribution of the GBIF records we have retrieved. As you can see below, most
of our data is recent, from after the year 2000. This means that our previous
results are likely to represent the current state of biodiversity
(considering a temporal autocorrelation effect). However, we agree that extinct
populations could be a real problem for the validation step of our method, and
therefore we decided to add a temporal restriction to our query. We have then
re-run our analyses with only GBIF occurrences registered after the year 2000.
Still, our results are overall the same. We have updated the absolute values on
lines 212-215, 225-228, 259, 274-275, and 285-289.

@@

![](figures/GBIF_years_prop_nb_2000.png)

We hope our revised manuscript is now more appropriate for publishing at PeerJ.
Reiterating our appreciation for yours and the reviewers' thoughtful comments,
we are available to answer any further questions.

### 2022-11-28 draft.tex
Deploying to gh-pages from @ graciellehigino/ms_range_interactions@e2706f4558e88565b6a05ee9c0ed8967a5f7c59c 🚀
Deploying to gh-pages from @ graciellehigino/ms_range_interactions@095143c312750503d13a65306e709771a9f87a1b 🚀

\vfill
\textbf{\sffamily Abstract: }\textbf{Background.} Range maps are a
useful tool to describe the spatial distribution of species. However,
they need to be used with caution, as they essentially represent a rough
approximation of a species' suitable habitats. When stacked together,
the resulting communities in each grid cell may not always be realistic,
especially when species interactions are taken into account. Here we
show the extent of the mismatch between range maps, provided by the
International Union for Conservation of Nature (IUCN), and species
interactions data. More precisely, we show that local networks built
from those stacked range maps often yield unrealistic communities, where
species of higher trophic levels are completely disconnected from
primary producers. \textbf{Methodology.} We used the well-described
Serengeti food web of mammals and plants as our case study, and provide
updated range maps for all predators by taking into account food-web
structure. We then used occurrence data from the Global Biodiversity
Information Facility (GBIF) to investigate where data is most lacking.
\textbf{Results.} We found that most predator ranges comprised large
areas without any overlapping distribution of their preys. However, many
of these areas contained GBIF occurrences of the predator.
\textbf{Conclusions.} Our results suggest that the mismatch between both
data sources could be due either to the lack of information about
ecological interactions or the geographical occurrence of preys. We
finally discuss general guidelines to help identify defective data among
distributions and interactions data, and we recommend this method as a
valuable way to assess whether the occurrence data that are being used,
even if incomplete, are ecologically accurate.
\vfill

@@

We indentified areas of data deficits on the ranges of predators based
on a simple rule: we removed any part of a predator's range that did not
intersect with the range of at least one prey herbivore species, which
in turn is directly connected to a primary producer (plants). To do
that, we used a Serengeti food web dataset (Baskerville et al. 2011)
(which comprises carnivores, herbivores, and plants from Tanzania) and
its species ranges from IUCN. Then, we calculated the difference in
range sizes between the original IUCN ranges of predators and those
without the areas where they would be alone, based on species
interaction data. Finally, we added the GBIF occurrence points for the
Serengeti species to investigate whether the results would be different
if we used another source of distribution data. \#\# Data

>>>>

\vfill
\textbf{\sffamily Abstract: }\textbf{Background.} Range maps are a
useful tool to describe the spatial distribution of species. However,
they need to be used with caution, as they essentially represent a rough
approximation of a species' suitable habitats. When stacked together,
the resulting communities in each grid cell may not always be realistic,
especially when species interactions are taken into account. Here we
show the extent of the mismatch between range maps, provided by the
International Union for Conservation of Nature (IUCN), and species
interactions data. More precisely, we show that local networks built
from those stacked range maps often yield unrealistic communities, where
species of higher trophic levels are completely disconnected from
primary producers. \textbf{Methodology.} We used the well-described
Serengeti food web of mammals and plants as our case study, and provide
updated range maps for all predators by taking into account food web
structure. We then used occurrence data from the Global Biodiversity
Information Facility (GBIF) to investigate where data is most lacking.
\textbf{Results.} We found that most predator ranges comprised large
areas without any overlapping distribution of their preys. However, many
of these areas contained GBIF occurrences of the predator.
\textbf{Conclusions.} Our results suggest that the mismatch between both
data sources could be due either to the lack of information about
ecological interactions or the geographical occurrence of preys. We
finally discuss general guidelines to help identify defective data among
distributions and interactions data, and we recommend this method as a
valuable way to assess whether the occurrence data that are being used,
even if incomplete, are ecologically accurate.
\vfill

@@

We indentified areas of data deficits on the ranges of predators based
on a simple rule: we removed any part of a predator's range that did not
intersect with the range of at least one prey herbivore species, which
in turn is directly connected to a primary producer (plants). To do
that, we used a Serengeti food web dataset (Baskerville et al. 2011)
(which comprises carnivores, herbivores, and plants from Tanzania) and
its species ranges from IUCN. Then, we calculated the difference in
range sizes between the original IUCN ranges of predators and those
without the areas where they would be alone, based on species
interaction data. Finally, we added the GBIF occurrence points for the
Serengeti species to investigate whether the results would be different
if we used another source of distribution data.

\hypertarget{data}{%
\subsection{Data}\label{data}}

### 2022-11-28 preprint.tex
Deploying to gh-pages from @ graciellehigino/ms_range_interactions@e2706f4558e88565b6a05ee9c0ed8967a5f7c59c 🚀
Deploying to gh-pages from @ graciellehigino/ms_range_interactions@095143c312750503d13a65306e709771a9f87a1b 🚀

\textbf{Abstract}:\,\textbf{Background.} Range maps are a useful tool to
describe the spatial distribution of species. However, they need to be
used with caution, as they essentially represent a rough approximation
of a species' suitable habitats. When stacked together, the resulting
communities in each grid cell may not always be realistic, especially
when species interactions are taken into account. Here we show the
extent of the mismatch between range maps, provided by the International
Union for Conservation of Nature (IUCN), and species interactions data.
More precisely, we show that local networks built from those stacked
range maps often yield unrealistic communities, where species of higher
trophic levels are completely disconnected from primary producers.
\textbf{Methodology.} We used the well-described Serengeti food web of
mammals and plants as our case study, and provide updated range maps for
all predators by taking into account food-web structure. We then used
occurrence data from the Global Biodiversity Information Facility (GBIF)
to investigate where data is most lacking. \textbf{Results.} We found
that most predator ranges comprised large areas without any overlapping
distribution of their preys. However, many of these areas contained GBIF
occurrences of the predator. \textbf{Conclusions.} Our results suggest
that the mismatch between both data sources could be due either to the
lack of information about ecological interactions or the geographical
occurrence of preys. We finally discuss general guidelines to help
identify defective data among distributions and interactions data, and
we recommend this method as a valuable way to assess whether the
occurrence data that are being used, even if incomplete, are
ecologically accurate.

>>>>

\textbf{Abstract}:\,\textbf{Background.} Range maps are a useful tool to
describe the spatial distribution of species. However, they need to be
used with caution, as they essentially represent a rough approximation
of a species' suitable habitats. When stacked together, the resulting
communities in each grid cell may not always be realistic, especially
when species interactions are taken into account. Here we show the
extent of the mismatch between range maps, provided by the International
Union for Conservation of Nature (IUCN), and species interactions data.
More precisely, we show that local networks built from those stacked
range maps often yield unrealistic communities, where species of higher
trophic levels are completely disconnected from primary producers.
\textbf{Methodology.} We used the well-described Serengeti food web of
mammals and plants as our case study, and provide updated range maps for
all predators by taking into account food web structure. We then used
occurrence data from the Global Biodiversity Information Facility (GBIF)
to investigate where data is most lacking. \textbf{Results.} We found
that most predator ranges comprised large areas without any overlapping
distribution of their preys. However, many of these areas contained GBIF
occurrences of the predator. \textbf{Conclusions.} Our results suggest
that the mismatch between both data sources could be due either to the
lack of information about ecological interactions or the geographical
occurrence of preys. We finally discuss general guidelines to help
identify defective data among distributions and interactions data, and
we recommend this method as a valuable way to assess whether the
occurrence data that are being used, even if incomplete, are
ecologically accurate.
